Reproducible Research : Peer Reviewed Assignment 1
==================================================

###Loading and preprocessing the data
1) Load the data
```{r,echo=FALSE}
library(data.table)
library(ggplot2)
if (!file.exists("activity.csv"))
{
        stop("activity.csv doesn't exist in working directory") 
}
```
```{r,echo=TRUE}
activity <- fread(file="activity.csv",sep = ",",na.strings = "NA",stringsAsFactors = FALSE,header=TRUE)
```
```{r,echo=FALSE}
activity$date <- as.Date(activity$date,format = "%Y-%m-%d")
```

2) In activity data frame, change 'date` column from character type to Date type
```{r,echo=TRUE}
activity$date <- as.Date(activity$date,format = "%Y-%m-%d")
```

###What is mean total number of steps taken per day?

1) Histogram of Total number of steps taken each day
```{r,echo=FALSE}
totalSteps_day <- data.table(aggregate(activity$steps,by=list(activity$date),FUN=sum))
```

```{r,echo=TRUE}
hist(totalSteps_day[!is.na(totalSteps_day$x),]$x, xlab="Total Steps taken per day", main="Histogram of Total Number of Steps taken per day", breaks=55)
abline(v=mean(totalSteps_day$x,na.rm = TRUE),col="green",lwd=2)
abline(v=median(totalSteps_day$x,na.rm = TRUE),col="blue",lwd=2)
```

2) Mean and Median of total number of steps taken
```{r, echo=FALSE}
mn <- mean(totalSteps_day$x,na.rm = TRUE)
mdn <-median(totalSteps_day$x,na.rm = TRUE)
```
        Mean   : `r as.integer(mn)`
        Median : `r mdn`

###What is the average daily activity pattern?

1) Time series plot 5 minute Interval Vs average number of steps taken averaged across all days

```{r,echo=FALSE}
activity_wtNA <- activity[activity[,!is.na(activity$steps)],]
avg_steps_interval <- data.table(aggregate(activity_wtNA$steps,by=list(activity_wtNA$interval),FUN=mean))
names(avg_steps_interval) <- c("Interval","AvgSteps")
```
```{r,echo=TRUE}
with(avg_steps_interval,plot(Interval,AvgSteps,typ="l",xlab="Interval", ylab="Average number of steps", col="blue", main="Time Interval Vs Average number of steps taken, averaged for all days"))
peak_avg <- avg_steps_interval[which(avg_steps_interval$AvgSteps==max(avg_steps_interval$AvgSteps)),]$Interval
abline(v=peak_avg,lwd=2,col="green")
```

2) The 5 min interval, on average across all the days in the dataset, contains maximum number of steps 

        Interval where average number of steps maximum  :  `r peak_avg`
        Average Max number of steps taken               :  `r max(avg_steps_interval$AvgSteps)`

###Imputing missing values
1) Total number of missing values (NA's) in the dataset
```{r,echo=FALSE}
missingVal <- nrow(activity_NA <- activity[activity[,is.na(activity$steps)],])
```
        Total number of rows having missing values : `r missingVal`


```{r,echo=FALSE}
avg_steps_day <- data.table(aggregate(activity$steps,by=list(activity$date),FUN=mean))
names(avg_steps_day) <- c("date","avgsteps")

#fill median values in NA in avg_steps_day
avg_steps_day[is.na(avg_steps_day$avgsteps),]$avgsteps <- median(avg_steps_day$avgsteps,na.rm=TRUE)

activity_imputing <- data.table(activity)

#activity_imputing[is.na(steps),]$steps <-  as.integer(avg_steps_day[activity_imputing[is.na(activity_imputing$steps),], on="date"]$avgsteps)

#fill NAs for respective days based on NAs in the activity_imputing dataset 
activity_imputing[is.na(steps),]$steps <-  as.integer(avg_steps_day[activity_imputing[is.na(steps),], on="date"]$avgsteps)

#histogram after imputing the NAs with median details
total_steps_day <- data.table(aggregate(activity_imputing$steps,by=list(activity_imputing$date),FUN="sum"))
names(total_steps_day) <- c("date","avgSteps")
```
2) Histogram of total number of steps taken

        Missing values are filled with median of average number of steps taken each day - `r as.integer(median(avg_steps_day$avgsteps,na.rm=TRUE))`)
        
```{r,echo=TRUE}
hist(total_steps_day$avgSteps,xlab="Total Number of Steps taken per day", main="Histogram of Total Number of Steps taken per day",breaks=61)
abline(v=mean(total_steps_day$avgSteps),col="green",lwd=2)
abline(v=median(total_steps_day$avgSteps),col="blue",lwd=2)
```

3) Mean and Median of total number of steps taken
```{r, echo=FALSE}
mn1 <- mean(total_steps_day$avgSteps)
mdn1 <-median(total_steps_day$avgSteps)
```
        Mean   : `r as.integer(mn1)`
        Median : `r mdn1` 

The Mean and Median values after imputing are slightly lower than previous computation of omitting the missing values. The impact to original dataset is low.

###Are there differences in activity patterns between weekdays and weekends?
```{r,echo=FALSE}
activity_imputing$dayOfWeek <- weekdays(activity_imputing$date)
activity_imputing[(dayOfWeek=="Sunday" | dayOfWeek=="Saturday"),]$dayOfWeek<-"Weekend"
activity_imputing[dayOfWeek!="Weekend",]$dayOfWeek<-"Weekday"
imput_avg_interval <- data.table(aggregate(activity_imputing$steps,by=list(activity_imputing$dayOfWeek,activity_imputing$interval),FUN="mean"))
names(imput_avg_interval) <- c("DayType","Interval","avgSteps")
imput_avg_interval <- transform(imput_avg_interval,DayType=factor(DayType,levels=c("Weekday","Weekend"),labels = c("Weekday","Weekend")))
```

1) Plot to show the difference in activity between Weekday and Weekend
        
        There seems to be higher level of activity during Weekend than Weekday
        
```{r,echo=TRUE,fig.width=8}
library(ggplot2)
ggplot(imput_avg_interval,aes(Interval,avgSteps))+geom_line(aes(col=DayType))+facet_wrap(~DayType, ncol=1)+labs(x="Interval",y="Average Number of steps", title="Interval Vs Average Number of steps in an interval, averaged by weekend and weekday")+geom_smooth(method = "lm",aes(col=DayType),se=FALSE)
```






